\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
first cutting pass

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are better optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution commonly performed in modern processors and propose what
we call speculative evaluation in PSO.  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

@@

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms, but
in all cases the number of processors never exceeds the number of particles.
The number of iterations of PSO that the algorithms can perform is thus
inherently limited by the time it takes to evaluate the objective
function---additional processors add more particles, but do not make the
iterations go any faster.

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
effect on the value obtained.  However, increasing the swarm size from 16 to 30
makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  Other functions show similar trends, though
often the optimal swarm size is slightly larger.  For this reason, previous
work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

In this paper we consider PSO parallelization strategies for clusters of
thousands of processors and functions for which a single evaluation will take
at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
is to explore the question of what to do with a thousand processors when 50 or
100 particles is the recommended swarm size, and simply increasing the swarm
size encounters diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) techniques commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-2002-constricted-pso}.

@@

The refactoring hinges on the idea that once the random coefficients are
determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
For the Ring topology there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update: a minus sign ($-$) represents no update, $L$ represents
an update to $\nbest$ coming from the left neighbor, $R$ represents an update
to $\nbest$ coming from the right neighbor, and $S$ represents an update to
either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
$\casexn$ refers to the case that the particle finds a new personal best, but
neither it nor its neighbors found a position that updated its neighborhood
best.  In the equations that follow, we refer to an unspecified update case as
$\casegen$, and to the set of cases collectively as $\caseset$.

@@

In order to incorporate the determination of which case occurs into the
position and velocity update equations, we introduce an indicator function
$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
$\casegen$ corresponds to the case actually taken by PSO,
$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
then sum over all of the cases, and the indicator function will make all of the
terms drop to zero except for the case that actually occurs.  For example, the
indicator function for the specific case $\casexn$ can be written as follows:

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

For each case $\casegen \in \caseset$, there is also a corresponding
velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
iteration $t$:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
			\ppos_{t}) \bigr]
\end{align}

In the same way we can create notation for the position update function by
substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
subscripts.
\begin{align}
\label{eq:defpcasegen}
	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
\end{align}

With this notation we can re-write the original PSO velocity
equation~\eqref{eq:velupdate}, introducing our sum over cases with the
indicator functions.  Again, we represent the indicator functions and velocity
functions without the parameters for compactness.  The equation becomes:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr]  \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
\end{align}

Similarly, the position update equation~\eqref{eq:posupdate} becomes:
\begin{align}
\nonumber
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
\label{eq:pos2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
\end{align}

@@

Returning our attention to the computation of $\ppos_{t+1}$ in
\eqref{eq:pos2update} and writing it with the parameters which were omitted
above, we obtain:
\begin{align}
\nonumber
  \ppos_{t+1} &= \sum_{c \in \caseset} \\
\nonumber
	& \indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
\label{eq:val2updatelong}
	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
\end{align}

In this form the important point to notice is that there are only 7 values (for
this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$ (note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration).  Thus PSO has been refactored
such that the algorithm can begin computing all 7 of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.

Although this speculative algorithm computes
$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
ignored, and might just as well never have been computed.  We call the set of
computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
``speculative children'' because only one of them will be needed.

To see the value of this refactoring, suppose that 1000 processors are
available, and that the evaluation of the function to be optimized takes one
hour.  If we only want a swarm of 100 particles, 900 of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
iterations in one hour.

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If we were not performing speculative
evaluation, those processors might very well be used to add particles to the
swarm.  This then raises the question of whether a swarm of 100 particles doing
twice as many iterations outperforms a swarm of 800 particles.  We show in the
rest of this paper that in many, though not all, instances, a smaller swarm
performing more iterations does in fact outperform a larger swarm.

@@

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to
effectively simulate the Complete topology.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology and
the Random topology on a number of common benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  In the rest of this
section we move from the mathematics of our method to its implementation.
First we discuss the relatively easy case of a centralized parallel PSO
algorithm with a master computer and many slaves.  In such an architecture, the
master keeps track of all necessary information with only trivial message
passing needed.  Then we discuss the more complicated case of a distributed
algorithm, where each particle is on its own and needs to send and receive
messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

@@

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
neighbor of particle $\p_t$.  Sets of particles are given by $\pset$, $\sset$,
or $\nset$, whereas single particles are simply $\p$, $\s$, or $\n$.

@@

\subsection{Centralized Algorithms}

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
speculative evaluation in such an architecture, the master generates the
positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
the slaves evaluate the objective function at all necessary positions, the
master then decides which position to accept for each particle, as in
\eqref{eq:val2updatelong}.  The outline of the procedure is given in
\alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  \eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$ according to
	  \eqref{eq:defpcasegen}.
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$ according to
	  \eqref{eq:val2updatelong}.
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using \eqref{eq:velupdate} and
\eqref{eq:posupdate} to obtain the set $\noeval{\pset}_t$.  For each particle
$\noeval{\p}_t$, the master must then get its set of neighbors
$\noeval{\nset}_t$ and use their positions, along with the position of
$\noeval{\p}_t$, to calculate all possible values of
$\specpos_{t+1}^{\casegen}$, using \eqref{eq:defpcasegen}.  These positions,
along with the original particle's associated information (such as values for
$\pbest$ and $\nbest$), define a set of speculative children,
$\noeval{\sset}_{t+1}$.  The master than has a set of particles
$\noeval{\pset}_t$, and for each particle a set of speculative children
$\noeval{\sset}_{t+1}$, which can all be evaluated at once.

The master then has the slaves evaluate the particles.  Once all particles,
speculative and original, have been evaluated and the values reported to the
master, the master determines which speculative child of each particle was the
correct one.  Mathematically, this corresponds to the evaluation of an
indicator function similar to that found in \eqref{eq:deficasexn}.  In
practice, this is done first by updating each particle's $\pbest$, if
necessary, then by updating the particle's $\nbest$ with information from the
particle's neighbors.  This is simply the original PSO algorithm, and
corresponds to steps 1--5 in \alg{centralized}.  Given the updates to $\pbest$
and $\nbest$, the case from Table~\ref{tab:evals} can be determined, as per
\eqref{eq:val2updatelong}.  The child with the matching case is kept, and all
other speculative children are discarded (step 7 in \alg{centralized}.

The parent $\p_t$ must pass its personal best value to the child, as the child
knows only the position that it guessed, not the function value at that
position.  It is possible that both $\p_t$ and $\s_{t+1}$ update their personal
bests, but $\p_t$'s value is better.  For example, suppose that $\p_{t-1}$ has
a personal best value of 3, and the we are seeking to minimize the function.
$\noeval{\p}_t$ is created, and $\noeval{\s}_{t+1}$ is moved assuming that
$\p_t$ has updated its personal best with its position at time $t$.  Then both
$\noeval{\p}_t$ and $\noeval{\s}_{t+1}$ are evaluated, with values 1 and 2,
respectively.  $\nonbest{\s}_{t+1}$ would think that its current position is
its personal best, as the value it found, 2, is better than its previous
personal best value of 3.  It needs to receive the personal best value from its
parent to know that its personal best position $\pbest$ is actually the
position of $\p_t$, not $\s_{t+1}$.

The parent also needs to pass the value of the neighborhood best that the child
guessed.  The child only knows the position and needs the value in order to
make future comparisons between neighborhood best positions (step 8).

Upon picking the correct branch for each particle and updating the child's
personal best and neighborhood best value (from iteration $t$), the result is
the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
speculative.  What remains is to update the neighborhood best of those
particles from their neighbors (from iteration $t+1$), as above, to obtain
$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$ (steps 1 and 2 in
\alg{centralized}), and the process repeats itself.

\subsection{Distributed Algorithms}

\label{sec:distributed}

>>>>

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are more efficiently optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution commonly performed in modern processors and propose what
we call speculative evaluation in PSO (SEPSO).  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

@@

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms, but
in all cases the number of processors does not exceed the number of particles.
The number of iterations of PSO that the algorithms can perform is thus
inherently limited by the time it takes to evaluate the objective
function---additional processors add more particles, but do not make the
iterations go any faster.

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
effect on the value obtained.  However, increasing the swarm size from 16 to 30
makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  Other functions show similar trends, though
often the optimal swarm size is larger.  For this reason, previous
work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

In this paper we consider PSO parallelization strategies for clusters of
hundreds of processors and functions for which a single evaluation will take
at least seconds, but probably minutes.  Our purpose
is to explore the question of what to do with a hundred processors when 50 or
100 particles is the recommended swarm size, and simply increasing the swarm
size encounters diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) a technique commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-2002-constricted-pso}.

@@

The refactoring hinges on the idea that once the random coefficients $\pURand_{t}$ and $\pURand_{t}$ are
determined, there are only a few possible updates to $\nbest$ and $\pbest$.
For the Ring topology there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update: a minus sign ($-$) represents no update, $L$ represents
an update to $\nbest$ coming from the left neighbor, $R$ represents an update
to $\nbest$ coming from the right neighbor, and $S$ represents an update to
either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
$\casexn$ refers to the case that the particle finds a new personal best, but
neither it nor its neighbors found a position that updated its neighborhood
best.  In the equations that follow, we refer to an unspecified update case as
$\casegen$, and to the set of cases collectively as $\caseset$.

@@

In order to incorporate the determination of which case occurs into the
position and velocity update equations, we introduce an indicator function
$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
$\casegen$ corresponds to the case taken by PSO,
$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
then sum over all of the cases, and the indicator function will make all of the
terms drop to zero except for the case that actually occurs.  For example, the
indicator function for the specific case $\casexn$ can be written as follows:

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ [ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) ]
	    \wedge [ \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) ] \\
	   &\quad \wedge \quad [ \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) ]
	    \wedge [ \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) ] \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

For each case $\casegen \in \caseset$, there is also a corresponding
velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
iteration $t$:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\label{eq:defvcasexn}
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
			\ppos_{t}) \bigr]
\end{align}

In the same way we can create notation for the position update function by
substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
subscripts.
\begin{align}
\label{eq:defpcasegen}
	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t})
	 = \ppos_{t} + \specvel_{t+1}^{\casegen}
\end{align}

With this notation we can re-write the original PSO velocity
equation~\eqref{eq:velupdate}, introducing our sum over cases with the
indicator functions.  Again, we represent the indicator functions and velocity
functions without the parameters for compactness.  The equation becomes:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr]  \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
\end{align}

Similarly, the position update equation~\eqref{eq:posupdate} becomes:
\begin{align}
\label{eq:pos2update}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
	= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
\end{align}

@@

Returning our attention to the computation of $\ppos_{t+1}$ in
\eqref{eq:pos2update} and writing it with the parameters which were omitted
above, we obtain:
\begin{align}
\nonumber
  \ppos_{t+1} &= \sum_{c \in \caseset} \bigl[
	\indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
\label{eq:val2updatelong}
	& \quad \quad \quad \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t}) \bigr]
\end{align}

In this form the important point to notice is that there are only $7$ values (for
this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$. Note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration.  Thus PSO has been refactored
such that the algorithm can begin computing all $7$ of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.

Although this speculative approach computes
$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
ignored, and might just as well never have been computed.  We call the set of
computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
``speculative children'' because only one of them will be needed.

To see the value of this refactoring, suppose that $800$ processors are
available, and that the evaluation of the function to be optimized takes one
hour.  If we only want a swarm of $100$ particles, $700$ of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
$\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle with $7$ speculative
evaluations ($700$ processors dedicated to speculative evaluation), thus using all $800$
processors and performing two
iterations in one hour.

In order to do two iterations at once, we use 8 times as many processors
as there are particles in the swarm.  If these processors were not performing speculative
evaluation, they might instead be used for function evaluation needed to support a large swarm.
This raises the question of whether a swarm of 100 particles doing
twice as many iterations outperforms a swarm of 800 particles.  We show in the
rest of this paper that in many, instances, a smaller swarm
performing more iterations does in fact outperform a larger swarm.
We acknowledge that both intuition and prior research~\cite{mcnabb-2009-large-particle-swarms}
indicate that the optimization of deceptive functions benefits greatly from
large and even very large swarm sizes. Thus this work will focus on less deceptive functions.

\subsection{Topology}
\label{sec:topology}

@@

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  In such cases, we use the Random topology 
which has been shown to
effectively approximate the rapid information flow of the Complete topology.
In our implementation of the random topology there are to randomly chosen neighbors
(chosen anew at every iteration)~\cite{mcnabb-2009-large-particle-swarms}.

In the rest of this
section we move from the mathematics of our method to its implementation.
First we discuss the relatively easy case of a centralized parallel PSO
algorithm with a master computer and many slaves.  In such an architecture, the
master keeps track of all necessary information with only trivial message
passing needed.  Then we discuss the more complicated case of a distributed
algorithm, where each particle is on its own and needs to send and receive
messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

@@

To aid in describing our implementation, we introduce a few terms. We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
neighbor of particle $\p_t$.  Sets of particles are given by $\pset$, $\sset$,
or $\nset$, whereas single particles are simply $\p$, $\s$, or $\n$.

@@

\subsection{Implementation}

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
speculative evaluation in such an architecture, the master generates the
positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
the slaves evaluate the objective function at all necessary positions, the
master then decides which position to accept for each particle, as in
\eqref{eq:val2updatelong}.  The outline of the speculative evaluation in PSO (SEPSO) algorithm is given in
\alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation PSO (SEPSO)}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  \eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$ according to
	  \eqref{eq:defpcasegen}.
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$ according to
	  \eqref{eq:val2updatelong}.
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

In a distributed parallel PSO implementation, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those
particles.  Messages such as
values and positions for the neighborhood best are sent between processors.
There may still be some machine that collects information from all of the
particles and outputs the result of the algorithm, though that machine's
importance is much less than in centralized algorithms.
In some distributed frameworks rounds of message passing can be expensive.
A N\:aive implementation of speculative evaluation
uses three rounds of message passing for every two iterations
(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
perform speculative evaluation in PSO with only one round of communication per
two iterations. 
Dynamic topologies like the random topology discussed above, further complicate the
needed message passing.
However, the message passing implementation is not particularly
interesting, but is tedious thus, in the interest of space, is omitted.
